Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-12-12 paper/abstract.tex
adding period gap section

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
proxy, to reveal their rotational evolution and demonstrate that kinematics
could be a useful tool for calibrating gyrochronology.
% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
% to the Praesepe cluster, accurately predicts the relative ages of stars.
We find that the period-color relationship of the Praesepe cluster is not
repeated in stars of older ages: a polynomial gyrochronology model, fit to
Praesepe stars, under-predicts the ages of late K dwarfs in the field.
In general we find that lower-mass stars spin more slowly than higher-mass
stars at young ages.
This is likely because their stronger magnetic fields lead to more efficient
angular momentum loss and reflects the behavior of young open clusters such as
the Pleaides, Praesepe, and the Hyades.
However, at older ages we find that late G and early K dwarfs rotate at the
{\it same rate} or faster than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results align with recent findings from the rotation periods of stars in
middle-aged open clusters and theoretical models that vary the rate of
surface-to-core angular momentum transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
an age proxy, to reveal their rotational evolution and demonstrate that
kinematics could be a useful tool for calibrating gyrochronology.
Firstly, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.
% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
% to the Praesepe cluster, accurately predicts the relative ages of stars.
We find that the period-color relationship of the Praesepe cluster is not
repeated in stars of older ages: a polynomial gyrochronology model, fit to
Praesepe stars, under-predicts the ages of late K dwarfs in the field.
In general we find that lower-mass stars spin more slowly than higher-mass
stars at young ages.
This is probably because their stronger magnetic fields lead to more efficient
angular momentum loss and reflects the behavior of young open clusters such as
the Pleaides, Praesepe, and the Hyades.
However, at older ages we find that late G and early K dwarfs rotate at the
{\it same rate} or faster than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results align with recent findings from the rotation periods of stars in
middle-aged open clusters and theoretical models that vary the rate of
surface-to-core angular momentum transport as a function of time and mass.
We also find that the kinematically oldest \kepler\ stars with measured
rotation periods are late-K and early-M dwarfs, indicating that these stars
maintain spotted surfaces and stay magnetically active longer than more
massive stars.
Finally, we find that the most rapidly rotating K-dwarfs are likely to be
synchronized binaries.

### 2019-12-12 paper/conclusion.tex
adding period gap section

We examined the rotational evolution of \kepler\ field stars using their
velocity dispersions as a proxy for age.
We found that old groups of cool dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This implies that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time in a
way that qualitatively agrees with theoretical that which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer, allowing their
rotation periods to be measured at older ages.
% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

>>>>

In this paper, we demonstrated that the dispersion of velocities in the
direction of galactic latitude, \vb, can be used as an age proxy, by showing
that there is no strong evidence for mass-dependent heating in low-mass
\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
increase with decreasing mass.
Although {\it vertical} velocity, \vz, is the quantity that most strongly
traces time-dependent orbital heating in the disc of the Galaxy, most stars
with measured rotation periods do not yet have radial velocities, so we used
velocity in the direction of Galactic latitude, \vb\, as a proxy for \vz.
Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
\vz\ introduces an additional velocity dispersion, which increases with
increasing Galactic latitude.
For this reason we did not attempt to convert \vb\ dispersions into ages using
an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore assumed that \vb\ velocity dispersion can be used to accurately
rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
on average, older than a group of stars with a small velocity dispersion.

% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
the evolution of stellar rotation period as a function of effective
temperature and age.
We found that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-\teff\ relation for old stars.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We showed that the period-\teff\ relations change shape over time in a way
that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.

% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
We also found that the oldest stars in the \mct\ catalog are cooler than 4500
K, which suggests that lower-mass stars remain active for longer, allowing
their rotation periods to be measured at older ages.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
The velocity dispersions of stars increase smoothly across the rotation period
gap, indicating that the gap does not separate two distinct populations.
Finally, we used kinematics to reveal a population of synchronized binaries
with rotation periods less than around 10 days.

% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

### 2019-12-12 paper/results.tex
adding period gap section

% To explore the relationship between rotation period, \teff\ and age for field
% K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
% ($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
% We used the \citet{angus2019} gyrochronology relation which is a simple,
% separable relation in \gcolor\ color and age, calibrated using the
% period-color relation of Praesepe and the period-age relation of Praesepe and
% the Sun.
% The large number of Praesepe members with precise rotation periods from the
% \ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
% types F through early M, makes it a good cluster for calibrating the
% period-color relation of stars at young ages.
% However, this is an extremely simple model and although it accurately
% describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
% of NGC 6819 -- the oldest cluster with available rotation periods), it
% over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
% cluster.
% These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
% figure \ref{fig:age_gradient}.
% The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
% along the upper edge of the MS, where stellar evolution models predict old
% stars to be, however the majority of these `old' stars are bluer than \gcolor\
% $\sim$ 1.5 dex.
% This suggests that either old M dwarfs are missing from the \mct\ catalog, or
% the \citet{angus2019} gyrochronology relation under-predicts the ages of
% low-mass stars.
% Given that lower-mass stars stay active for longer than higher-mass stars
% \citep[\eg][]{west2011}, and are therefore more likely to have measurable
% rotation periods at old ages, the latter scenario seems more likely.

% To further explore the relationship between rotation period, \teff\ and age,
% we first removed high and low velocity outliers by performing 3$\sigma$
% sigma-clipping on the \vb\ velocities.
% Without sigma-clipping, we found that a small number of high velocity
% outliers at the low-temperature end of our sample substantially raised the
% velocity dispersion for cooler stars.
% We then selected groups of stars within different gyrochronal age ranges and
% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
% of effective temperature for each age group.
% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
% throughout this paper we show rotation periods as a function of effective
% temperature, \teff, because it is easier to divide stars into bins of roughly
% equal numbers in \teff-space than in color-space.

@@

To explore the relationship between rotation period, \teff\ and velocity
dispersion, we first removed high and low velocity outliers from the \mct\
sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remain the
same with, or without sigma-clipping.
We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
3500 K to avoid biases caused by the selection function at the faint, cool
end, and binarity or weakened braking \citep{vansaders2016} at the hot end.

Figure \ref{fig:vplot} shows rotation period versus effective temperature for
the \mct\ sample, coloured by the standard deviation of their (\vb)
velocities, where \sigmavb\ was calculated for groups of stars over a grid in
$\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show isochrones that project the rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These isochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages: they do not trace lines of constant
    velocity dispersion.
    Isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from the Kepler Input Catalog][]{brown2011}.
    White lines show isochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these isochrones appear to reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:main_figure}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that both velocity
dispersion and rotation period increases with age as expected.
Lines of constant age (isochrones) appear to follow the shape of the
Praesepe-based gyrochronology model (black dotted line) at young ages.
However, at older ages it appears that the relation between rotation period
and \teff\ flattens out, until eventually rotation period {\it decreases} with
decreasing effective temperature at a given age.

Figure \ref{fig:masses_with_models} shows the same as figure \ref{fig:vplot},
however mass is plotted on the x-axis instead of temperature and the models
are those of \citet{spada2019}.
The masses of these stars are taken from the \kepler\ input catalog
\citep{brown2011}.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts over a narrow range of temperatures.

These results indicate that stars of spectral type ranging from late G to late
K follow a braking law that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.

The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass, likely because
lower-mass stars with deeper convection zones have stronger magnetic fields,
larger Alfv\'en radii and therefore experience greater angular momentum loss
rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for GK dwarfs.
The explanation for this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two interior layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly.

@@

% % The groups of stars with the largest velocity dispersions are cooler than 4500
% % K, presumably because cooler stars remain active for longer, and their
% % rotation periods can therefore be measured at older ages.

@@

% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
% are faster than expected; their rotational evolution appears to have stalled,
% and they have similar rotation periods to the 650 Myr Praesepe cluster
% \citep{curtis2019}.
% The rotation periods of the K dwarfs in this cluster are plotted in figure
% \ref{fig:dispersion_period_teff}.
% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
% K dwarfs lie below it.
% This cluster may be the `missing link' that connects two epoch of stellar
% spin-down.
% % , however without more observations of middle-aged open clusters it
% % : an early stage where the period-\teff\ relation for cool dwarfs has
% % a negative slope and a late stage where it has a positive slope.
% % In this case the period gap may delineate the transition between these two
% % regimes and is the point at which stellar magnetic dynamos likely undergo a
% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

>>>>

% To explore the relationship between rotation period, \teff\ and age for field
% K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
% ($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
% We used the \citet{angus2019} gyrochronology relation which is a simple,
% separable relation in \gcolor\ color and age, calibrated using the
% period-color relation of Praesepe and the period-age relation of Praesepe and
% the Sun.
% The large number of Praesepe members with precise rotation periods from the
% \ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
% types F through early M, makes it a good cluster for calibrating the
% period-color relation of stars at young ages.
% However, this is an extremely simple model and although it accurately
% describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
% of NGC 6819 -- the oldest cluster with available rotation periods), it
% over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
% cluster.

% We then selected groups of stars within different gyrochronal age ranges and
% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
% of effective temperature for each age group.
% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
% throughout this paper we show rotation periods as a function of effective
% temperature, \teff, because it is easier to divide stars into bins of roughly
% equal numbers in \teff-space than in color-space.

@@

To explore the relationship between rotation period, \teff\ and velocity
dispersion, we first removed high and low velocity outliers from the \mct\
sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remain the
same with, or without sigma-clipping.
We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
3500 K to avoid biases caused by the selection function at the faint, cool
end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with short rotation periods by cutting out stars
with gyrochronal ages less than 0.5 Gyr.

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show isochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These isochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages: they do not trace lines of constant
    velocity dispersion.
    Isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from the Kepler Input Catalog][]{brown2011}.
    White lines show isochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these isochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that both velocity
dispersion and rotation period increases with age as expected.
Black lines show isochrones from the \citep{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These isochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
At the youngest ages, these isochrones seem to describe the data well: the
palest yellow (youngest) stars with the lowest velocity dispersions all fall
close to the 0.5 Gyr isochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr isochrones do appear to trace constant
velocity dispersion/age among the field stars, by 1.5 Gyrs the isochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr isochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The isochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones however, they should follow lines of constant
velocity dispersion.
At ages older than around 1 Gyr, it appears that isochrones should have a more
flattened, or even inverted relationship between rotation period and effective
temperature than these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, obtained from the Kepler Input
Catalog \citep{brown2011}, with isochrones from the \citep{spada2019}
gyrochronology model shown as white lines.
These isochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
Although these models do not trace lines of constant velocity dispersion at
0.5 and 1 Gyr, they do appear to reproduce trends in the data at ages of 1.5
Gyr and older.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts.

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K follow a braking law that changes over
time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass, likely because
lower-mass stars with deeper convection zones have stronger magnetic fields,
larger Alfv\'en radii and therefore experience greater angular momentum loss
rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for K-dwarfs.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

It has been demonstrated that lower-mass stars remain magnetically active
longer than more massive stars, \citep{west2008, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.

\subsection{The period gap and synchronized binaries}
\label{sec:gap}

There is a sharp gap in the population of rotation periods, which lies just
above the 1 Gyr isochrone in the upper panel of figure \ref{fig:vplot}, whose
origin is unknown and is the subject of much speculation \citep{mcquillan2014,
davenport2018, reinhold2019}.
The Praesepe-based model appears to be valid below the gap but not above.
Although this may be coincidental (and more data would be needed to confirm a
connection) the gap may indeed separate a young regime where stellar cores are
decoupled from their envelopes from an old regime where these layers are more
tightly coupled.
If so, this could indicate that the phenomenon responsible for changing the
shape of isochrones in rotation-\teff\ space is related to the phenomenon that
produces the gap.

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperate for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Only stars cooler than 5000 K are plotted in
    the bottom panel in order to isolate populations above and below the
    period gap, which only extends up to temperatures of $\sim$4600 K.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

Figure \ref{fig:gap} shows the velocity dispersions of stars in the \mct\
sample, with stars subdivided into three groups: those that rotate more
quickly than the major rotation period distribution (green triangles), those
with rotation periods shorter than the gap (orange squares), and those with
rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology models, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
The velocity dispersion of stars with rotation periods shorter than the lower
edge of the rotation period distribution (green triangles) is relatively
large.
This is an indication that many of these are synchronized binaries.
The rotation periods of stars in synchronized binaries are tidally locked to
the orbital period of the binary system and since synchronized binaries have
short orbital periods this can produce old, seemingly isolated stars, with
short rotation periods.
The large velocity dispersions of the most rapidly rotating stars indicates
that some fraction of these stars are old but rotating rapidly, and therefore
likely to be synchronized binaries.
Figure \ref{fig:gap} indicates that there is an increased probability of stars
with rotation periods less than $\sim$10 days being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of visual binaries were rapid rotators, and the probability of a star
being a synchronized binary system substantially increased at rotation periods
of around 7 days \citep{simonian2019}.

% We calculated the overall velocity dispersions of stars as a function of
% rotation period and gyrochronal age.

@@

@@

% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
% are faster than expected; their rotational evolution appears to have stalled,
% and they have similar rotation periods to the 650 Myr Praesepe cluster
% \citep{curtis2019}.
% The rotation periods of the K dwarfs in this cluster are plotted in figure
% \ref{fig:dispersion_period_teff}.
% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
% K dwarfs lie below it.
% This cluster may be the `missing link' that connects two epoch of stellar
% spin-down.
% % , however without more observations of middle-aged open clusters it
% % : an early stage where the period-\teff\ relation for cool dwarfs has
% % a negative slope and a late stage where it has a positive slope.
% % In this case the period gap may delineate the transition between these two
% % regimes and is the point at which stellar magnetic dynamos likely undergo a
% \subsection{Velocity dispersion of coeval groups}
% \label{sec:age_cut}
% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.
